---
title: "Trafford Assist"
output: html_document
---

```{r include=FALSE}
#   DEFINING THE LIBRARIES 

library(tidyverse) ; library(sf) ; library(leaflet) ; library(htmltools)
library(dplyr); library(janitor); library(leaflet.extras) ; library(htmlwidgets); 
library(tseries); library(ggplot2); library(plotly); library(zoo)

```

```{r include=FALSE}

# -------  COLLECTING THE DATA ---------

# collecting 2016/17 client dataset
df_client_16_17 <- read.csv('TraffordAssist_ClientDetails_2016_2017.csv')
df_client_16_17$year <- '2016/2017'

# collecting 2017/18 client dataset
df_client_17_18 <- read.csv('TraffordAssist_ClientDetails_2017_2018.csv')
df_client_17_18$year <- '2017/2018'

# collecting 2016/17 furniture dataset
df_furniture_16_17 <- read.csv('TraffordAssist_Furniture_2016_2017.csv', colClasses=c(Application.Reference = 'character'), fileEncoding="latin1")

# collecting 2017/18 furniture dataset
df_furniture_17_18 <- read.csv('TraffordAssist_Furniture_2017_2018.csv', colClasses=c(Application.Reference = 'character'), fileEncoding="latin1")

# collecting 2016/17 food dataset
#Application.Reference is converted to character, the same data type as Application.Reference within df_client
#in order to perform the matching using left_join
df_food_16_17 <- read.csv('TraffordAssist_Food_2016_2017.csv', colClasses=c(Application.Reference = 'character'), fileEncoding="latin1")

# collecting 2017/18 food dataset
df_food_17_18 <- read.csv('TraffordAssist_Food_2017_2018.csv', colClasses=c(Application.Reference = 'character'), fileEncoding="latin1")

# collecting 2016/17 pay point dataset
df_paypoint_16_17 <- read.csv('TraffordAssist_PayPoint_2016_2017.csv', colClasses=c(Application.Reference = 'character'), fileEncoding="latin1")

# collecting 2017/18 paypoint dataset
df_paypoint_17_18 <- read.csv('TraffordAssist_PayPoint_2017_2018.csv', colClasses=c(Application.Reference = 'character'), fileEncoding="latin1")

# load postcode data (Source: ONS Postcode Directory)
postcodes <- read_csv("trafford_postcodes_2018-11.csv")

```

```{r include=FALSE}

# -------- ASSIGNING CATEGORIES ------------

df_furniture_16_17$category <- 'furniture'
df_furniture_17_18$category <- 'furniture'
df_food_16_17$category <- 'food'
df_food_17_18$category <- 'food'
df_paypoint_16_17$category <- 'paypoint'
df_paypoint_17_18$category <- 'paypoint'


```

```{r include=FALSE}

# ------------   SELECTING RELEVANT COLUMNS FROM THE DATA ----------------

# select a subset of columns from the 2016/17 client dataset
df_client_16_17 <- df_client_16_17 %>% select(Application.Reference, Customer.Name, Caseworker, Age, Gender,
                                              Postcode, Date.application.received..dd.mm.yyyy., 
                                              Date.of.Decision..dd.mm.yyyy., Month, year)


colnames(df_client_16_17) <- c('application_ref', 'customer_name', 'user', 'age', 'gender', 'postcode', 
                               'application_received_date', 'application_processed_date', 'month', 'year')

# select a subset of columns from the 2017/18 client dataset
df_client_17_18  <- df_client_17_18  %>% select(Application.Reference, Customer.Name, User, Age, X,
                                                Postcode, App.Date, Processed.Date, Month, year)


colnames(df_client_17_18) <- c('application_ref', 'customer_name', 'user', 'age', 'gender', 'postcode', 
                               'application_received_date', 'application_processed_date', 'month', 'year')


# select a subset of columns from the 2016/17 furniture dataset
df_furniture_16_17 <- df_furniture_16_17 %>% select(Application.Reference, category)
colnames(df_furniture_16_17) <- c('application_ref', 'category')

# select a subset of columns from the 2017/18 furniture dataset
df_furniture_17_18 <- df_furniture_17_18 %>% select(Application.Reference, category)
colnames(df_furniture_17_18) <- c('application_ref', 'category')

# select a subset of columns from the 2016/17 food dataset
df_food_16_17 <- df_food_16_17 %>% select(Application.Reference, category)
colnames(df_food_16_17) <- c('application_ref', 'category')

# select a subset of columns from the 2017/18 food dataset
df_food_17_18 <- df_food_17_18 %>% select(Application.Reference, category)
colnames(df_food_17_18) <- c('application_ref', 'category')

# select a subset of columns from the 2016/17 pay point dataset
df_paypoint_16_17 <- df_paypoint_16_17 %>% select(Application.Reference, category)
colnames(df_paypoint_16_17) <- c('application_ref', 'category')

# select a subset of columns from the 2017/18 pay point dataset
df_paypoint_17_18 <- df_paypoint_17_18 %>% select(Application.Reference, category)
colnames(df_paypoint_17_18) <- c('application_ref', 'category')


```

```{r include=FALSE}

# ----------- MERGING DATASETS ------------

# combinding 2016/17 AND 2017/18 client dataset
df_client <- rbind(df_client_16_17, df_client_17_18)

# combinding 2016/17 AND 2017/18 furniture dataset and remove NA
df_furniture <- rbind(df_furniture_16_17, df_furniture_17_18)
df_furniture <- df_furniture[!is.na(df_furniture$application_ref), ]

# combinding  2016/17 and 2017/18 food dataset and remove NA
df_food <- rbind(df_food_16_17, df_food_17_18)
df_food <- df_food[!is.na(df_food$application_ref), ]

# combinding  2016/17 and 2017/18 paypoint dataset and remove NA
df_paypoint <- rbind(df_paypoint_16_17, df_paypoint_17_18)
df_paypoint <- df_paypoint[!is.na(df_paypoint$application_ref), ]

# combinding all category data
df_category <- rbind(df_furniture, df_food, df_paypoint)

# get just the unique categories
df_category <- unique(df_category)

# joining client dataset with furniture, food and paypoint datasets to obtain the category
df <- unique(df_client)
df <- left_join(df, df_category, by = "application_ref")


```

```{r include=FALSE}

# ----------- LINKING WITH POSTCODE DATASET AND FILTERING TRAFFORD ASSIST DATASET TO CREATE THE FINAL DATASETS -------------


# clean the postcode from the dataset
df$postcode <- toupper(df$postcode)
df$postcode <- gsub("/FURN", "", df$postcode)

# match postcodes to retrieve coordinates
geo <- left_join(df, postcodes, by = "postcode")

# setting up the 'WHERE' condition to filter the dataset
geomatch_condition <- !is.na(geo$area_name)
geo_match_found <- geo[which(geomatch_condition), ]
geo_no_match_found <- geo[-which(geomatch_condition),]

# further cleansing on the geo dataset 
df_final <- geo[which(geomatch_condition | geo$postcode == 'NFA'), ]
df_final <- df_final[which(!is.na(df_final$category)), ]

#dataframe filtered by furniture
df_final_furniture <- df_final[df_final$category == 'furniture', ]

#dataframe filtered by food
df_final_food <- df_final[df_final$category == 'food', ]

#dataframe filtered by paypoint
df_final_paypoint <- df_final[df_final$category == 'paypoint', ]

```

```{r include=FALSE}

# ------- PREPARING DATASETS FOR CREATING CHOROPLETH MAP -------------

# convert to spatial object 
sf <- df_final %>% st_as_sf(crs = 4326, coords = c("lon", "lat"), na.fail=FALSE)

# building  dataframes of ward name and number of applications per ward, for each categories
wards <- levels(factor(df_final$area_name))

wards_df_all <- data.frame(matrix(ncol = 2))
wards_df_furniture <- data.frame(matrix(ncol = 2))
wards_df_food <- data.frame(matrix(ncol = 2))
wards_df_paypoint <- data.frame(matrix(ncol = 2))

colnames(wards_df_all) <- c('ward_name', 'num_of_applications')
colnames(wards_df_furniture) <- c('ward_name', 'num_of_applications')
colnames(wards_df_food) <- c('ward_name', 'num_of_applications')
colnames(wards_df_paypoint) <- c('ward_name', 'num_of_applications')

i = 1
for(w in wards){
  wards_df_all[i,1] <- w
  wards_df_furniture[i,1] <- w
  wards_df_food[i,1] <- w
  wards_df_paypoint[i,1] <- w
  
  wards_df_all[i, 2] <- nrow(df_final[which(!is.na(df_final$area_name) & df_final$area_name == w),])
  wards_df_furniture[i, 2] <- nrow(df_final_furniture[which(!is.na(df_final_furniture$area_name) & df_final_furniture$area_name == w),])
  wards_df_food[i, 2] <- nrow(df_final_food[which(!is.na(df_final_food$area_name) & df_final_food$area_name == w),])
  wards_df_paypoint[i, 2] <- nrow(df_final_paypoint[which(!is.na(df_final_paypoint$area_name) & df_final_paypoint$area_name == w),])
  
  i = i + 1
}


# load the ward data for great britain
sf_gb <- st_read("https://opendata.arcgis.com/datasets/07194e4507ae491488471c84b23a90f2_3.geojson", quiet = TRUE)

# filter to retrieve only the wards in Trafford
lookup <- read_csv("https://opendata.arcgis.com/datasets/046394602a6b415e9fe4039083ef300e_0.csv") %>%
  filter(LAD17NM %in% c("Trafford")) %>%
  pull(WD17CD)

sf_trafford <- sf_gb %>% filter(wd17cd %in% lookup)
#plot(st_geometry(sf_trafford))

names(sf_trafford)[2] <- 'area_code'
names(sf_trafford)[3] <- 'ward_name'
names(sf_trafford)[7] <- 'lon'

# merge Trafford Assist data (containing the ward name and the number of applications per ward) with Trafford Spatial data
sf_trafford_all <- left_join(sf_trafford, wards_df_all, by = "ward_name")
sf_trafford_furniture <- left_join(sf_trafford, wards_df_furniture, by = "ward_name")
sf_trafford_food <- left_join(sf_trafford, wards_df_food, by = "ward_name")
sf_trafford_paypoint <- left_join(sf_trafford, wards_df_paypoint, by = "ward_name")

```

```{r include=FALSE}

# ------- CREATING THE CHOROPLETH MAP -------------

map <- leaflet() %>% 
  setView(-2.35533522781156, 53.419025498197, zoom = 12) %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolylines(data = sf_trafford_all, stroke = TRUE, weight = 2, color = "#212121", opacity = 1) %>%
  addPolygons(data = sf_trafford_all, group = 'All', fillColor = ~colorQuantile("RdPu", domain = num_of_applications, probs = seq(0, 1, length.out = 8))(num_of_applications), weight = 2, opacity = 1, color = "white", dashArray = "3", fillOpacity = 0.7,
               highlight = highlightOptions(weight = 3, color = "#000", dashArray = "", fillOpacity = 0.7, bringToFront = TRUE),
               label = sprintf("<strong>%s</strong><br/>%g applications", sf_trafford_all$ward_name, sf_trafford_all$num_of_applications ) %>% lapply(htmltools::HTML), 
               labelOptions = labelOptions(style = list("font-weight" = "normal", padding = "3px 8px"), textsize = "15px", direction = "auto")) %>%
  addPolygons(data = sf_trafford_furniture, group = 'Furniture', fillColor = ~colorQuantile("RdPu", domain = num_of_applications, probs = seq(0, 1, length.out = 8))(num_of_applications), weight = 2, opacity = 1, color = "white", dashArray = "3", fillOpacity = 0.7,
              highlight = highlightOptions(weight = 3, color = "#000", dashArray = "", fillOpacity = 0.7, bringToFront = TRUE),
              label = sprintf("<strong>%s</strong><br/>%g applications", sf_trafford_furniture$ward_name, sf_trafford_furniture$num_of_applications ) %>% lapply(htmltools::HTML), 
              labelOptions = labelOptions(style = list("font-weight" = "normal", padding = "3px 8px"), textsize = "15px", direction = "auto")) %>%
  addPolygons(data = sf_trafford_food, group = 'Food', fillColor = ~colorQuantile("RdPu", domain = num_of_applications, probs = seq(0, 1, length.out = 8))(num_of_applications), weight = 2, opacity = 1, color = "white", dashArray = "3", fillOpacity = 0.7,
              highlight = highlightOptions(weight = 3, color = "#000", dashArray = "", fillOpacity = 0.7, bringToFront = TRUE),
              label = sprintf("<strong>%s</strong><br/>%g applications", sf_trafford_food$ward_name, sf_trafford_food$num_of_applications ) %>% lapply(htmltools::HTML), 
              labelOptions = labelOptions(style = list("font-weight" = "normal", padding = "3px 8px"), textsize = "15px", direction = "auto")) %>%
  addPolygons(data = sf_trafford_paypoint, group = 'Paypoint', fillColor = ~colorQuantile("RdPu", domain = num_of_applications, probs = seq(0, 1, length.out = 8))(num_of_applications), weight = 2, opacity = 1, color = "white", dashArray = "3", fillOpacity = 0.7,
              highlight = highlightOptions(weight = 3, color = "#000", dashArray = "", fillOpacity = 0.7, bringToFront = TRUE),
              label = sprintf("<strong>%s</strong><br/>%g applications", sf_trafford_paypoint$ward_name, sf_trafford_paypoint$num_of_applications ) %>% lapply(htmltools::HTML), 
              labelOptions = labelOptions(style = list("font-weight" = "normal", padding = "3px 8px"), textsize = "15px", direction = "auto")) %>%
  addLegend(data = sf_trafford_all, pal = colorQuantile("RdPu", domain = sf_trafford_all$num_of_applications, probs = seq(0, 1, length.out = 8)), 
            values = ~sf_trafford_all$num_of_applications, opacity = 0.7, title = NULL, position = "bottomright") 
  
map <- map %>% 
  addLayersControl(baseGroups = c("All", "Furniture", "Food", "Paypoint"),
                   #overlayGroups = tiers,
                   options = layersControlOptions(collapsed = FALSE),
                   position = "bottomleft")
```

```{r include=FALSE}

# ---------- PREPARING DATASET FOR TIME SERIES LINE GRAPH

time_df <- data.frame(matrix(ncol = 3))
colnames(time_df) <- c('year', 'month', 'num_of_applications')

years <- c('2016', '2017', '2018')
months <- c(4,5,6,7,8,9,10,11,12,1,2,3)
for(m in months){
  date <- as.Date(df_final$application_received_date,'%d/%m/%Y')
  month <- format(date,'%m')
  year <- format(date,'%Y')
  
}

# create a dataframe of the all dates for an application, and create a thrid column with a frequency of 1
df_date <- as.Date(df_final$application_received_date,'%d/%m/%Y')
d_df <- data.frame(format(date,'%Y'), format(date,'%m'), as.integer(1))
colnames(d_df) <- c('year', 'month', 'freq')

# group the dates by year (Group.1) and month (Group.2) and apply a sum function 
# to determine the number of applications for each month in each year
d_df_agg = aggregate(d_df$freq, by = list(d_df$year, d_df$month), FUN = sum)
d_df_agg <- d_df_agg[order(d_df_agg$Group.1, d_df_agg$Group.2),]
d_df_agg <- d_df_agg[5:28,]

# convert the dataset into a time series data
d_ts <- ts(d_df_agg$x, start = c(2016, 4), end = c(2018, 3), frequency = 12)

# convert the time series data into a dataframe to be used by ggplot function
d_df_ts <- data.frame(y=d_df_agg$x, x=time(d_ts), year=d_df_agg$Group.1, month=d_df_agg$Group.2, month_year=as.yearmon(paste(d_df_agg$Group.1, d_df_agg$Group.2, sep = "-")))

```

```{r include = FALSE}

# ---------------- CREATING THE TIME SERIES LINE GRAPH ------------

#pl <- plot(d_ts, xlab='year', ylab='frequency')

pl <- ggplot(d_df_ts, aes(as.Date(x), y)) + 
  geom_line() + 
 labs(x='year', y='No of applications', title='Plot of number of applications between April 2016 and March 2018')

ts_tooltip = paste("Number of applications: ", d_df_ts$y, "\n", "Date: ", d_df_ts$month_year )

pp <- plotly_build(pl)
ts_graph <- style( pp, text=ts_tooltip, hoverinfo = "text")

```


```{r include = FALSE}

# -------------- CREATING BAR PLOTS -----------------

# bar plots - food
pl_b_fo <- ggplot(wards_df_food, aes(x=ward_name, y=num_of_applications, fill=as.factor(wards_df_food$ward_name))) + 
  geom_bar(stat='identity') + 
  theme_bw() +
  theme(axis.text.x=element_blank(), legend.position = 'none')  +
  labs(x='wards', y='No of applications', title='Plot of number of food applications per ward for 2016/17 - 2017/18', fill= NULL)

b_fo_graph <- ggplotly(pl_b_fo, tooltip = c('ward_name', 'num_of_applications'))

# bar plots - furniture
pl_b_fu <- ggplot(wards_df_furniture, aes(x=ward_name, y=num_of_applications, fill=as.factor(wards_df_furniture$ward_name))) + 
  geom_bar(stat='identity') + 
  theme_bw() +
  theme(axis.text.x=element_blank(), legend.position = 'none')  +
  labs(x='wards', y='No of applications', title='Plot of number of furniture applications per ward for 2016/17 - 2017/18', fill=NULL)

b_fu_graph <- ggplotly(pl_b_fu, tooltip = c('ward_name', 'num_of_applications'))

# bar plots - paypoint
pl_b_pp <- ggplot(wards_df_paypoint, aes(x=ward_name, y=num_of_applications, fill=as.factor(wards_df_paypoint$ward_name))) + 
  geom_bar(stat='identity') + 
  theme_bw() +
  theme(axis.text.x=element_blank(), legend.position = 'none')  +
  labs(x='wards', y='No of applications', title='Plot of number of pay point applications per ward for 2016/17 - 2017/18', fill=NULL)


b_pp_graph <- ggplotly(pl_b_pp, tooltip = c('ward_name', 'num_of_applications'))

# bar plots - all
pl_b_all <- ggplot(wards_df_all, aes(x=ward_name, y=num_of_applications, fill=as.factor(wards_df_paypoint$ward_name))) + 
  geom_bar(stat='identity') + 
  theme(axis.text.x=element_blank()) + 
  labs(x='wards', y='No of applications', title='Plot of number of applications per ward for 2016/17 - 2017/18', fill=NULL) +
  ylim(-100,700) +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank()
    #plot.margin = unit(rep(-2,4), "cm")     # This remove unnecessary margin around plot
  ) +
  coord_polar(start = 0) 


```

```{r, echo=FALSE}
# ------- DISPLAYING THE CHOROPLETH MAP -----------
map
```

<br/>

```{r, echo=FALSE}
# ------- DISPLAYING THE TIME SERIES LINE GRAPH -----------
ts_graph
```

<br/>

```{r, echo=FALSE}
# ------- DISPLAYING THE FOOD BAR GRAPH  -----------
b_fo_graph
```

<br/>

```{r, echo=FALSE}
# ------- DISPLAYING THE FURNITURE BAR GRAPH  -----------
b_fu_graph
```

<br/>

```{r, echo=FALSE}
# ------- DISPLAYING THE PAY POINT BAR GRAPH  -----------
b_pp_graph
```

<br/>

```{r, echo=FALSE}
# ------- DISPLAYING BAR GRAPH FOR ALL CATEGORIES -----------
pl_b_all
#b_all_graph
```

<br/>

```{r include = FALSE}

# ----------- GET THE SUMMARY STATISTICS

total_app <- nrow(df_final)
total_furniture <- nrow(df_final_furniture)
total_food <- nrow(df_final_food)
total_paypoint <- nrow(df_final_paypoint)

max_all <- wards_df_all[which.max(wards_df_all$num_of_applications),]
max_all_ward <- max_all$ward_name
max_all_num_app <- max_all$num_of_applications

max_furniture <- wards_df_furniture[which.max(wards_df_furniture$num_of_applications),]
max_furniture_ward <- max_furniture$ward_name
max_furniture_num_app <- max_furniture$num_of_applications 


max_food <- wards_df_food[which.max(wards_df_food$num_of_applications),]
max_food_ward <- max_food$ward_name
max_food_num_app <- max_food$num_of_applications

max_paypoint <- wards_df_paypoint[which.max(wards_df_paypoint$num_of_applications),]
max_paypoint_ward <- max_paypoint$ward_name
max_paypoint_num_app <- max_paypoint$num_of_applications

max_app_for_date <- d_df_ts[which.max(d_df_ts$y),]
max_ts_date <- max_app_for_date$month_year
max_ts_num_app <- max_app_for_date$y
```


<p>There are a total of **`r total_app `** applications made within Trafford for the years 2016/17 and 2017/18. Of these, **`r total_furniture `** are for furniture, **`r total_food `** are for food, and **`r total_paypoint`** are for pay point</p>

<p>There is a maximum of **`r max_furniture_num_app `** within **`r max_furniture_ward `** , for a <strong>furniture</strong> application</p>

<p>There is a maximum of **`r max_food_num_app `** within **`r max_food_ward `** , for a <strong>food</strong> application</p>

<p>There is a maximum of **`r max_paypoint_num_app `** within **`r max_paypoint_ward `** , for a <strong>pay point</strong> application</p>

<p>There is a maximum of **`r max_all_num_app `** within **`r max_all_ward `** , across <strong>all</strong> catgories</p>

<p>The most applications of **`r max_ts_num_app `** were made in **`r max_ts_date `**

